Manuscript draft

### 2024-04-15 manuscript.qmd
moved .qmd docs to root

```{r}
#| include: false
library(tidyverse)
library(here)
```

@@

An individual's social centrality may be determined by many factors. Differences in social position have been attributed to individual phenotypic characteristics, such as age, sex, or personality \[REFS\]. Of course, how an animal uses space is also an important driver of how many conspecifics it interacts with, and how frequently. The relationship between movement, space use, and social interaction patterns has been an area of active research in recent years. Webber et al [-@webber2023] have characterized the variation in patterns of animal movement and space use as the "spatial phenotype," as part of a larger framework for understanding the relationships between social and spatial drivers of animal behavior. In this light, the "spatial phenotype" may be considered in addition to more conventional phenotypic measures to explain why some individuals are more central than others in their social networks.

@@

Eurasian griffon vultures (*Gyps fulvus*) provide a study system well suited to ascertaining the relationship between individuals' spatial phenotypes and their social centrality.

-   obligate scavengers, large soaring fliers, lots of movement

-   Unusual GPS tracking coverage of the population–high proportion of individuals tracked, so we can get a population-level sense of movement behavior and social interaction patterns

-   Three clear social situations in which they interact, each with a different significance/relevance.

    -   Communal roosts: many individuals together at once, variability in size/density of roosts. Used for information exchange \[REFS\] and also important for disease transmission \[or not–cf Elvira Nili\]. Breeding/nesting also occurs at/near roost sites, on cliffs.

    -   Flight: interactions in flight are important for information transfer [@harel2017], finding locations of thermal uplift [@sassi2023] and feeding sites [@spiegel2013; @cortés-avizanda2014].

    -   Feeding: Communal feeding at carcasses. Also sometimes standing around the carcass site. Sometimes associate for a while before beginning to feed, e.g. waiting for critical mass to be reached (Orr Spiegel, pers. comm.) Potentially also important for disease transmission \[but not, cf. ref Elvira Nili\].

While some griffon vultures make long-range forays \[REFS\], there is a mostly resident population in a relatively small area in southern Israel \[REF\]. The population is small and restricted enough that individuals' home ranges are highly overlapping, with most individuals using a large fraction of available space in the region. This makes habitat features likely to affect most individuals similarly, though age has been shown to factor into flight skill and could therefore impact movement [@harel2016]. GPS fixes are taken once every 10 minutes, frequently enough to measure flight movements with good accuracy, and most individuals are tracked for several months to several years. Individuals have previously been shown to vary in their social network positions in each of the three social situations, with centrality in one situation not necessarily predicting centrality in another and each situation contributing differently to the overall social structure of the population [@sharma2023].\
\
\[General introduction to the importance of time scale in social network analysis\]

-   [@blonder2012; @cantor2012; @hobson2013; @rasmussen2013; @holme2015; @farine2018; @fisher2021; @pasquaretta2021]

Social situations occur over different time scales, and this is an important consideration when studying the causes and consequences of their structure. In defining co-roosting, co-flight, and co-feeding interactions based on proximity in each of these behavioral situations, we have created social networks in which edges already conceptually represent interactions over different amounts of time. Because individuals have one roosting location per night, edge weights in the co-roosting network are measured in units of days, while pairs of individuals can interact many times in flight or at a carcass on a given day.\
\
The biological processes driving the execution of these behaviors and the formation of these networks are likewise happening over different time scales. An individual must decide once per day which roost to go to [@harel2016]. But flight is a continuous process driven by many decisions resulting from individual and social information [@sassi2023], and feeding events are concentrated in time on a given day. Feeding events are driven by the temporal rhythm of the appearance of carcasses, which occur naturally or are provisioned every \[NUMBER\] days and persist for hours to a few days (Orr Spiegel, pers. comm.). Individuals may return to the same carcass on successive days. While at a carcass, individuals may fight and interact many times.

# Research Questions

@@

Q3: How does the time scale of movement predict social interactions?

-   H3A: The relationship between movement and social centrality will be strongest within the time scales at which change occurs in the social network.

    -   P3A: Co-flight networks will change hourly to daily, co-feeding networks will change every few days, and co-roosting networks will change over longer periods of time (weeks to months).

    -   P3B: Social position in each of the social situations will be best predicted by individuals' movement patterns aggregated over the corresponding time scale.

# Methods

## Study system

To answer these questions, we analyzed data from a free-flying population of Eurasian griffon vultures (Gyps fulvus) fitted with GPS tags.

@@

Over the period \[date\] to \[date\], the dataset contained \[number\] of tagged vultures. Tags recorded GPS fixes, as well as speed and heading readings, every 10 minutes when battery conditions allowed. The GPS data are stored on Movebank [@kays2022] and were loaded into R using the `move` package\[ @kranstauber2023\].

## Tagging and aging

## Data cleaning

We subsetted the dataset to exclude periods of time that vultures spent in capture cages, in the wildlife hospital following poisonings or injuries, or in captivity prior to wild release (in the case of captive-born birds). We cleaned the data to remove invalid GPS fixes. Then we split data from each into three seasons. The fall, or "pre-breeding," season includes dates from 15 September through 14 December; the breeding season lasts from 15 December through 14 May of the following year; and the summer lasts from 15 May through 14 September. These season boundaries were chosen based on biological knowledge of the vultures' breeding timelines (see more details in the Study System section). Therefore, the analysis contained a total of 8 seasons, from the fall of 2020 through the 2023 breeding season (Table X). All of these data was reserved for use in proximity-based social network analysis. Since the GPS tags give few to no fixes overnight, and since reception is often very poor on the cliffs where vultures roost, we combined information from individuals' last locations of the evening and first locations of the morning to assign a roost site for each individual on each night \[more details\]. Full details about the data cleaning and roost assignment procedure can be found in the supplementary material.

To characterize individual movement behavior and spatial phenotype, we cleaned the data further. We restricted movement points to the period between sunrise and sunset (calculated for each day with respect to a point at the centroid of Israel using the R package suncalc [@thieurmel2022]. We restricted movement calculations to a set of focal individuals with activity regions centered in the Negev region of Israel, the site of most of our tagging effort. For each vulture, we also removed days with too few GPS fixes. Fixes were ideally collected every 10 minutes, but sometimes they were less frequent due to bad cellular reception in the cliff areas where vultures often sit or due to low battery charge. For each individual, only days with 10 or more GPS fixes per day were retained, unless the battery charge never dipped below 50% during that day. In those cases, the low number of points was attributed to the vultures spending most of their time sitting on a cliff without much movement, rather than to missed points due to low battery. Since these sedentary days represent an important component of vulture movement behavior, we retained this information (Orr Spiegel, pers. obs.).

\[Geography\]

For each season, we calculated movement measures only for individuals tracked for at least 30 days (after removing low-point days as described above). We calculated flight altitude by extracting ground elevation information from elevation rasters obtained from the `elevatr` R package [@hollister2022], at a zoom level of \[level–more details about resolution?\] and subtracting those numbers from the vultures' tag-mounted altimeter readings. Finally, since some tags were occasionally set to a much higher fix rate (e.g. every 1 minute) for short periods of time, we downsampled the data to a 10 minute fix rate to make sure that all individuals had comparable fix rates.

## Movement calculations

For each focal individual, we characterized a pattern-based spatial phenotype (Webber et al. 2023) by calculating a variety of movement and space use measures over the course of each of the eight seasons. We calculated home range as the 95% AKDE of a vulture's GPS fixes in a season @

@@

We identified in-flight fixes (ground speed \>= 5 m/s) and calculated measures of daily flight behavior, including maximum and median daily flight altitude and daily flight time (the sum of time intervals between adjacent in-flight fixes). Finally, we used all fixes to calculate daily distance traveled (the sum of all distances between successive fixes); daily maximum displacement (the straight-line distance between the beginning of the daily movement track and the farthest point reached on that day); and movement tortuosity (cumulative distance to the point of maximum displacement divided by the straight-line distance to the point of maximum displacement [@benhamou2004]). We took seasonal means of each of these individual daily measures to get a single value per individual vulture, per season.

@@

To summarize individuals' spatial phenotype, we created three composite indices: **movement**, **space use**, and **roost diversification**. Each index was derived from a principal component analysis of several spatial phenotype variables, all of which were scaled and centered over the entire date range of the study before putting them into the PCA.

**Movement** was defined as the first principal component of a PCA containing per-vulture seasonal means of daily mean flight altitude, daily flight duration, daily maximum displacement, and daily flight tortuosity. **Space use** was defined as the first principal component of a PCA containing log-transformed 95% AKDE home range sizes and the ratio of the 50% AKDE to the 95% AKDE. **Roost diversification** was defined as the first principal component of a PCA containing a seasonal Shannon diversity index of roost sites used \[ref Shannon\], number of unique roost sites used, and the proportion of nights in a given season when a vulture switched to a new roost site (versus roosting at the same site as on the previous night). The composite measures accounted for %, %, and % of the variation in their respective PCAs.

@@

Co-feeding

Vultures that were not flying (ground speed \< 5 m/s, [@spiegel2013]) who came within 50m of each other for at least two consecutive 10-minute time periods, were considered to be having a ground interaction [@sharma2023]. Any ground interactions that occurred at night or within the boundaries of a known roost site were excluded. Therefore, the remaining diurnal ground interactions should represent vultures' interactions while feeding or while present at feeding sites (Orr Spiegel and Nili Anglister, pers. comm.). Co-roosting Roost sites were assigned using [methods]. Unlike flight and diurnal ground GPS fixes, vultures were only assigned one roosting location per night. Two vultures that roosted within the boundaries of the same roost site polygon on a given night were considered to be co-roosting.

@@

We used the R package `igraph` [@csardi2005] to calculate each individual's degree and strength for each social situation in each season. Degree is defined as the number of other individuals that a vulture is connected to (i.e. ever interacted with). Strength measures how much an individual interacts, and is calculated as the sum of the weights of all edges that the focal vulture is involved in.

@@

We modeled the relationship between the spatial phenotype variables and the network measures using linear mixed-effects regression models, fitted using the lme4 and lmerTest packages [@bates2015; @kuznetsova2017]. We fit separate models for each response variable (degree and strength). The predictors included were movement (PC1), exploration (PC2), situation (feeding, flight, roosting), age group (juvenile/subadult, adult), and season (breeding, summer, fall). To account for the different numbers of individuals in the network in each unique season, we included a random effect for the interaction of season and year, e.g. "2022_fall" (8 levels). To account for consistent individual differences in network position or behavior, and because individuals varied in how many seasons they were included in, we also included an individual-level random effect.

@@

```{r}
#| include: false
source(here("scripts/04.0_mixedModels.R"))
```

```{r}
#| echo: false
#| label: fig-d-space
#| fig-cap: "Relationship between normalized degree and space use"
plot_d_eff_space
```

```{r}
#| echo: false
#| label: fig-d-roost
#| fig-cap: "Relationship between normalized degree and roost diversification"
plot_d_eff_roost
```

```{r}
#| echo: false
#| label: fig-d-space-situ
#| fig-cap: "Relationship between normalized degree and space use, with separate trend lines for each social situation"
plot_d_eff_situ_space
plot_d_emt_situ_space
```

```{r}
#| echo: false
#| label: fig-s-movement
#| fig-cap: "Relationship between normalized strength and movement"
plot_s_eff_movement
```

```{r}
#| echo: false
#| label: fig-d-movement-situ
#| fig-cap: "Relationship between normalized degree and movement, with separate trend lines for each social situation"
plot_d_eff_situ_movement
plot_d_emt_situ_movement
```

```{r}
#| echo: false
#| label: fig-s-roost
#| fig-cap: "Relationship between normalized strength and roost diversification"
plot_s_eff_roost
```

```{r}
#| echo: false
#| label: fig-s-space
#| fig-cap: "Relationship between normalized strength and space use"
plot_s_eff_situ_space
plot_s_emt_situ_space
```

>>>>

```{r}
#| include: false
library(tidyverse)
library(here)
tar_config_set(store = here::here('_targets'))
```

#### Social network centrality is known to be important in animals.

@@

#### We observe that some individuals are more central than others in a social network. What causes this variation?

An individual's social centrality may be determined by many factors. Differences in social position have been attributed to individual phenotypic characteristics, such as age, sex, or personality \[REFS\].

##### Individual traits

-   Multiple spatial behaviors → Social centrality [@albery2021]

-   Exploration → Weak associations with more conspecifics, more movement between flocks. [@aplin2013]

-   Slow-exploring → More stable relationships [@aplin2013]

-   Personality → Social structure [@aplin2013]

-   Stress → Weaker parental bonds, in captivity [@boogert2014]

-   Stress → Foraging with more others, less choosy, more central, in captivity [@boogert2014]

-   Stress → Foraging with more others, less choosy, more central, in the wild [@brandl2019]

-   More exploratory → higher degree, in finches [@moyers2018]

-   Quick explorers → More interactions [@moyers2018]

-   Slower explorers → less central ("Yet, slower explorers overall did not travel shorter distances than faster explorers, indicating that a less central social network position was not merely a consequence of lower activity.") [@snijders2014]

-   Females → higher strength, more preferred partners (only during lactation) [@wey2013]

-   Females → more stable partnerships [@wey2013]

#### Individuals' spatial phenotypes are important determinants of their social centrality

Of course, how an animal uses space is also an important driver of how many conspecifics it interacts with, and how frequently. The relationship between movement, space use, and social interaction patterns has been an area of active research in recent years. Webber et al [-@webber2023] have characterized the variation in patterns of animal movement and space use as the "spatial phenotype," as part of a larger framework for understanding the relationships between social and spatial drivers of animal behavior. In this light, the "spatial phenotype" may be considered in addition to more conventional phenotypic measures to explain why some individuals are more central than others in their social networks.

-   Definition of "spatial phenotype" [@webber2023]

-   Definition of "spatial behaviour" [@albery2021]

-   Note that explicit spatial context is also important, and we didn't consider it. (We should have! How do we justify not doing that?)

    -   An individual's observed spatial phenotype results from a combination of its internal tendency toward movement and the external environmental conditions through which it moves [(Nathan et al. 2008)](https://www.zotero.org/google-docs/?IyWvRB)

#### Social interactions occur when individuals encounter each other, in various behavioral situations.

-   Co-occurrence in space and time is necessary for social interaction to occur. Therefore, more movement → more social interaction, in general.

-   However, movement is also an expression of an individual's underlying tendency to move, as well as potentially an emergent consequence of things like dominance rank or social environment.

-   Social centrality arises from a combination of the number of interactions (strength) and unique individuals encountered. An individual can achieve high social centrality by being more geographically central in the population, versus by being more versatile.

-   No one has investigated the link between spatial phenotype and social centrality. We know that individuals vary in both spatial phenotype and social centrality, but we don't know how the one influences the other.

    -   Furthermore, social networks are not monolithic. Because the biological mechanisms behind social network structure are different in different social situations, it stands to reason that spatial phenotype should be related to social centrality in different ways in the different social contexts.

        -   For example, different components of spatial phenotype might be expected to relate differently to networks in different social situations.

        -   Also, we might just see different-direction effects. For example, individuals that move a lot might interact more frequently in flight but interact less frequently at feeding stations (either because they spend less time feeding or because they feed at more peripheral locations). Therefore, it is important to consider different behavioral situations separately when teasing apart the relationship between the social and spatial phenotypes.

@@

#### We set out to uncover whether spatial phenotypes could explain observed variation in social centrality.

-   Hypotheses/predictions from slides

##### Research Questions

@@

# Methods {data-link="methods"}

### Study system

This work was conducted on a population of GPS-tagged Eurasian griffon vultures in southern Israel, a study system well suited to studying the relationship between movement and sociality. Griffon vultures are large obligate scavengers that use soaring flight to locate food resources (carcasses) on the landscape. They may travel long distances without expending very much energy. Griffon vultures are locally endangered in Israel, where they have declined significantly since \[DATE\], and they face significant anthropogenic threats, most commonly through pesticide and NSAID poisonings, power line collisions, and shootings. The population in southern Israel has an unusually high GPS tracking coverage, with a high proportion of individuals (approximately 80-90%) fitted with GOS tags. The tags send a position reading ever ten minutes if battery condition is good, or less frequently otherwise. \[paper\] says that we only need \[percentage\] of individuals in order to get a good sense of the population-level social structure, and our coverage far exceeds that. We are able to get a very good sense of individuals' movement behavior and social interaction patterns.

Griffon vultures interact in several social situations, each with a different biological relevance. In particular, griffons rely on social information to locate food resources. Interactions in flight (i.e. when individuals fly within each other's perceptual distance) provide an opportunity for local enhancement of carcass resources, or following to a carcass site. Therefore, they are important for information transfer [@harel2017], finding locations of thermal uplift [@sassi2023], and feeding sites [@spiegel2013; @cortés-avizanda2014]. Vultures also obtain social information about food resources at communal roosting sites. Roost sites can vary in size and density, with many vultures roosting together at once. These roosts also serve as breeding colonies during the breeding season (approximately December through May). Finally, vultures interact frequently at or around carcasses while eating. The free-living griffon vultures in the southern Israel population are provisioned at several feeding sites, where livestock carcasses are deposited irregularly to supplement the availability of naturally-occurring carcasses on the landscape. Vultures may co-feed, associate in close proximity near a carcass, or even fight over the food resource.

The southern Israel griffon vulture population is mostly resident, though some individuals (especially juveniles) occasionally make long-range forays \[REF\]. The population is small and restricted enough that individuals' home ranges are highly overlapping, with most individuals using a large fraction of available habitat in the region. This makes habitat features likely to affect many individuals similarly, though age has been shown to factor into flight skill and could therefore impact movement [@harel2016]. GPS fixes are taken frequently enough to monitor flight movements with good accuracy, and most individuals are tracked for several months to several years. Individuals have previously been shown to vary in their social network positions in each of the three social situations, with centrality in one situation not necessarily predicting centrality in another and each situation contributing differently to the overall social structure of the population [@sharma2023].

## Field methods

@@

## Data preparation

```{r}
tar_load(cleaned_r)
tar_load(joined0_r)
pctremovedcleaned <- (joined0_r[1]-cleaned_r[1])/joined0_r[1]
```

For this study, we used data between September 1, 2020 and September 15, 2023 from the \[ornitela\] and \[inpa\] datasets, which we downloaded from Movebank [@kays2022]. For each vulture, we removed any periods of time spent in the wildlife hospital (e.g. while recovering from an illness, injury, or poisoning), or in captivity prior to wild release (in the case of captive-born birds). We cleaned the data by filtering out GPS fixes deemed unreliable (e.g. too few satellites for triangulation, nonsensical heading readings, impossible jumps from previous fixes), and we removed any time that vultures spent in capture cages during the tagging season in the fall. Altogether, approximately `r pctremovedcleaned`% of fixes were removed as compared to the raw data. Next, we removed any vultures that spent fewer than 30 days, in total, within a focal region encompassing the general vicinity of Israel (in order to exclude, for instance, birds that were briefly detected in the area before dispersing far away). For vultures detected in the area for at least 30 days during the study period, we subsequently removed any fixes for those individuals that occurred outside of the same focal region, in order to restrict our analyses to movements that occurred near the main population locus and ensure that most social interactions would be with other tagged individuals. (We accounted for missing time periods, including removed data, by calculating association indices in the social networks).

We divided the remaining data into seasons based on biological knowledge of vultures' breeding timelines: a breeding season (December 15 through May 14 of the following year); a summer or post-breeding season (May 15th through September 14th), and a transient or pre-breeding season (September 15th through December 14th). For simplicity, we refer to each breeding season by the calendar year in which it ends (e.g. the 2021 breeding season lasts from December 15, 2020 through May 14, 2021). We analyzed data from the 2020 pre-breeding/transient season through the 2023 post-breeding/summer season, for a total of nine seasons across four calendar years (three of each type of season).

The vulture population in Israel is geographically separated, with most individuals in the south but a small and relatively isolated population at the Hai Bar Carmel Nature Reserve and in the Golan Heights. While vultures occasionally move between the northern and southern population, such movements are rare, and northern vultures behave differently than southern vultures due to more consistent feeding and a more restricted range in the north. Therefore, for each season, we restricted our analysis to individuals whose spatial centroid (including all GPS points after data cleaning) lay south of UTM Northing 3550000 (Zone 36N), equivalent to approximately 32.086238 North latitude. This threshold was determined by visual examination of a histogram of centroid latitudes, which was bimodal and disjoint. Finally, to account for older GPS tags that were set to lower fix frequencies, we excluded any individuals from each season that never had their tags set to a 10-minute frequency during that season. In recent months, some tags were temporarily set to record fixes at much higher frequencies, so those individuals were downsampled to a 10-minute frequency in order to make data analysis more tractable.

```{r}
tar_load(downsampled_10min_forSocial_r)
nvultures <- map_dbl(downsampled_10min_forSocial_r, "vultures")
```

After this cleaning, the data used for social network analysis contained between `r min(nvultures)` and `r max(nvultures)` vultures per season (mean = `r mean(nvultures)`; sd = `r sd(nvultures)`).

#### Determining roost locations

We used GPS fixes to determine vultures' nightly roost locations. The solar-powered GPS tags are set to record very few fixes overnight, and cell reception (needed for data transmission) is often very poor on the cliffs where roosting sites are located. Therefore, we combined information from individuals' last locations of the evening and first locations of the morning to assign a roost site for each individual on each night. Full details about the data cleaning and roost assignment procedure can be found in the supplementary material.

#### Further cleaning for movement analysis

To characterize individual movement behavior and spatial phenotype, we cleaned the data further. We restricted movement points to the period between sunrise and sunset (calculated for each day with respect to a point at the centroid of Israel using the R package {suncalc} [@thieurmel2022]. We restricted movement calculations to a set of focal individuals with activity regions centered in the Negev region of Israel, the site of most of our tagging effort. For each vulture, we also removed days with too few GPS fixes. Fixes were ideally collected every 10 minutes, but sometimes they were less frequent due to bad cellular reception in the cliff areas where vultures often sit or due to low battery charge. For each individual, only days with 10 or more GPS fixes per day were retained, unless the battery charge remained above 50% for the entire day. In those cases, the low number of points was attributed to the vultures spending most of their time sitting on a cliff (with limited reception) without much movement, rather than to actual missing data due to low battery. Since these sedentary days represent an important component of vulture movement behavior, we retained this information (Orr Spiegel, pers. obs.).

```{r}
tar_load(downsampled_10min_r)
nvultures <- map_dbl(downsampled_10min_r, "vultures")
```

After removing days with too few fixes per day, we removed any vultures that did not have at least 30 days of valid movement data remaining in a given season, to ensure that movement estimates and averages were based on a high enough sample size of days to capture the range of the vulture's behavior. Since this was done on a per-season basis, individuals that were removed from one season's analysis due to not having enough valid days could still have movement measures calculated for a different season. Finally, we downsampled the movement data to a 10-minute interval to remove high-frequency fixes. Ultimately, we calculated movement measures for a total of `r nvulturesmov` **focal individuals** in total (range: `r min(nvulturesmov_szns)`-`r max(nvulturesmov_szns)`; mean = `r round(mean(nvulturesmov_szns), 2)` vultures per season, sd = `r round(sd(nvulturesmov_szns), 2)` vultures per season).

## Measuring spatial phenotype

We characterized the pattern-based spatial phenotype [@webber2023a] of each focal individual by measuring its overall space use, use of roost sites, and movement patterns.

We calculated seasonal home ranges and core areas using autocorrelated kernel density estimates (AKDEs) via the {ctmm} R package [@fleming2023]. We defined the home range as the 95% AKDE and the core area as the 50% AKDE. To measure differences in how individuals used their home ranges, we calculated the ratio of the seasonal core area size to the seasonal home range size.

@@

We identified in-flight fixes (ground speed \>= 5 m/s) and calculated measures of daily flight behavior, including maximum and median daily flight altitude and daily flight time (the sum of time intervals between adjacent in-flight fixes). We calculated flight altitude by extracting ground elevation information from elevation rasters obtained from the {elevatr} package at a ground resolution of approximately 250 meters [@hollister2022] and subtracting those numbers from the vultures' tag-mounted altimeter readings. Finally, we used all fixes to calculate daily distance traveled (the sum of all distances between successive fixes); daily maximum displacement (the straight-line distance between the beginning of the daily movement track and the farthest point reached on that day); and movement tortuosity (cumulative distance to the point of maximum displacement divided by the straight-line distance to the point of maximum displacement [@benhamou2004]). We took seasonal means of each of these individual daily measures to get a single value per individual, per season.

@@

```{r}
#| include: false
load(here("data/calcMovement/pca_space.Rda"))
propvar_space <- as.data.frame(summary(pca_space)$importance)$PC1[2]

load(here("data/calcMovement/pca_roost.Rda"))
propvar_roost <- as.data.frame(summary(pca_roost)$importance)$PC1[2]

load(here("data/calcMovement/pca_movement.Rda"))
propvar_movement <- as.data.frame(summary(pca_movement)$importance)$PC1[2]
```

We characterized individuals' spatial phenotypes using three composite indices: movement, space use, and roost diversification. Each index was derived from a principal component analysis of several spatial phenotype variables, each scaled and centered over the entire date range of the study.

**Movement** was defined as the first principal component of a PCA containing per-vulture seasonal means of daily mean flight altitude, daily flight duration, daily maximum displacement, and daily flight tortuosity. It explained `r 100*round(propvar_movement,2)`% of the variation in the PCA. Individuals with high movement scores traveled farther and flew longer, with straighter daily trajectories overall. **Space use** was defined as the first principal component of a PCA containing log-transformed 95% AKDE home range sizes and the ratio of the 50% AKDE to the 95% AKDE. It explained `r 100*round(propvar_space,2)`% of the variation in the PCA. Individuals low on space use had small overall home ranges with concentrated core areas (and some exploratory movements away from the core), while individuals high on space use had large home ranges consisting of space that they used frequently (i.e. the core area made up a larger proportion of the home range). **Roost diversification** was defined as the first principal component of a PCA containing a seasonal Shannon diversity index of roost sites used \[ref Shannon\], number of unique roost sites used, and the proportion of nights in a given season when a vulture switched to a new roost site (versus roosting at the same site as on the previous night). It explained `r 100*round(propvar_roost,2)`% of the variation in the PCA.

While the griffon vulture population in southern Israel is fairly well mixed and vultures do not have discrete, exclusive territories, there is still variability in where individuals spend their time within the landscape. Spatial position has the potential to directly affect how many other individuals a vulture encounters and which feeding and roost sites it frequents. In addition, though a high proportion of the vultures in the area are GPS-tagged, more peripheral vultures might be more likely to interact with untagged birds living farther from the population center, potentially diluting their measured measured strength or degree. To account for the influence of geographic centrality on social centrality, we calculated the geographic centroid of all GPS points for each individual in each season. Then we compared these centroids to the overall centroid of all individuals in the population in that season to obtain an index of centrality (versus peripherality). We incorporated spatial centrality as a covariate into our linear models.

@@

### Co-feeding

Vultures that were not flying (ground speed \< 5 m/s, [@spiegel2013]) who came within 50m of each other for at least two consecutive 10-minute time periods, were considered to be having a ground interaction [@sharma2023]. Any ground interactions that occurred at night or within the boundaries of a known roost site were excluded. Therefore, the remaining diurnal ground interactions should represent vultures' interactions while feeding or while present at feeding sites (Orr Spiegel and Nili Anglister, pers. comm.).

### Co-roosting

Vultures were considered to be co-roosting if they roosted within 500m of each other on a given night. Unlike interactions in the co-feeding and co-flight networks, co-roosting interactions could occur at a maximum frequency of once per night. Therefore, edge weights in the co-roosting network represented the number of nights on which two individuals co-roosted divided by the number of nights when both individuals had a valid roost location (i.e. they theoretically could have co-roosted), regardless of proximity.

@@

We used the R package {igraph} [@csardi2005] to calculate each individual's degree and strength for each social situation in each season. Degree is defined as the number of other individuals that a vulture is connected to (i.e. ever interacted with). Strength measures how much an individual interacts, and is calculated as the sum of the weights of all edges that the focal vulture is involved in. We normalized degree by dividing it by the number of individuals present in the network in the current season minus one. We normalized strength by dividing it by the sum of all strength values in the network in the current season \[REF\].

## Comparison to a reference model

Since social interactions can arise from both spatial proximity and social preference (i.e. "passive" versus "active" interactions [@smith-aguilar2016]). We can use spatially explicit data-stream randomization to compare individuals' observed interaction rates to those expected to arise from chance encounters by individuals sharing space without synchronizing their movements [@gahm; @spiegel2016]. This allows us to decouple social and spatial drivers of social interactions. It also helps us account for how individuals' locations in space (e.g. the locations of their home ranges) affect how many social interactions they end up having.

Using the "wrap-around" method outlined in [@gahm], we conducted 100 iterations of the wrap-around randomization for each season of data separately, using a maximum time-shift of 1 day either backward or forward. We calculated normalized degree and strength for each individual in each season in each of the 100 replicate randomizations. Then, we compared each individual's observed normalized degree and strength to the distribution of 100 normalized degree and strength values obtained from the randomization replicates and calculated a z-score for the observed value to measure the extent to which it deviated from random. These z-scores were used as the response variables in our mixed models.

@@

We modeled the relationship between the spatial phenotype variables and the network measures using linear mixed-effects regression models, fitted using the {lme4} and {lmerTest} packages [@bates2015; @kuznetsova2017]. We fit separate models for each response variable (degree and strength). The predictors included were movement (PC1), exploration (PC2), situation (feeding, flight, roosting), age group (juvenile/subadult, adult), season (breeding, summer, fall), and geographic centrality. To account for the different numbers of individuals in the network in each unique season, we included a random effect for the interaction of season and year, e.g. "2022_fall" (8 levels). To account for consistent individual differences in network position or behavior, and because individuals varied in how many seasons they were included in, we also included an individual-level random effect.

@@

```{r}
#| include: false
#| output: false
source(here("scripts/04.0_mixedModels.R"))
```

## Deviations from random

Individual vultures interacted with significantly more unique others over the course of a season than expected by chance (i.e. observed degree values compared to the distribution of degree values obtained from networks created from the wrap-around data), across all seasons and social situations. However, individuals' strength did not differ from random.

```{r}
#| echo: false
tar_load(linked)

linked %>%
  select(seasonUnique, situ, z_deg, z_str) %>%
  ungroup() %>%
  pivot_longer(cols = contains("z_"), names_to = "measure", values_to = "z") %>%
  ggplot(aes(x = z, col = situ, group = interaction(seasonUnique, measure, situ)))+
  geom_density(aes(linetype = measure))+
  facet_wrap(~seasonUnique, scales = "free")+
  geom_vline(aes(xintercept = 0)) # notably, we see that the z_scores for degree are almost entirely positive, while the ones for strength are basically centered around zero, with very few exceptions.
```

## Degree

```{r}
#| label: fig-d-space
#| fig-cap: "Relationship between movement and degree deviation"
plot_d_eff_movement
```

```{r}
#| label: fig-d-space-situ
#| fig-cap: "Relationship between normalized degree and space use, with separate trend lines for each social situation"
plot_d_eff_situ_space
plot_d_emt_situ_space
```

There was a significant relationship between how vultures used space and how many other individuals they encountered in flight. Vultures that consistently ranged widely encountered more others in flight, while those that concentrated their movement in smaller home ranges and smaller proportional core areas flew with fewer others. Similarly, there was a marginally significant positive relationship between space use and degree deviation in the co-feeding network as well. By contrast, vultures with smaller home ranges/proportionally smaller core areas had a co-roosting degree that deviated more from chance than those with larger home ranges/proportionally larger core areas.

```{r}
#| label: fig-d-roost
#| fig-cap: "Relationship between normalized degree and roost diversification"
plot_d_eff_roost 
```

Vultures with greater roost diversification, had, overall, smaller degree deviations than those with smaller roost diversification, though the effect size was small.

## Strength
